Determinism, Complexity, and Predictability in Computer Performance

### 2014-03-17 model.tex
rewrite of ARIMA section and some notation changes there and in methods.tex

In this section, we describe the four different forecasting schema
used in this study, as well as the error metric used to evaluate their
predictive accuracy:
\begin{itemize}
\item The \emph{random-walk} method, which uses the previous value in
  the observed signal as the forecast,

@@

\item The \emph{LMA} method, which uses a near(est) neighbor forecast
  strategy on a dynamical reconstruction of the signal.
\end{itemize}
ARIMA is based on standard autoregressive techniques.  LMA is designed
to capture and exploit the deterministic structure of a signal from a
nonlinear dynamical system.  The first two methods, somewhat
surprisingly, often outperform sophisticated prediction strategies in
the case of highly complex signals, as discussed below.

@@

A random-walk predictor simply uses the last observed measurement as
the forecast: that is, the value $p(i)$ at time $i$ is predicted to
be $$p_i = X_{i-1,obs}$$ The \naive ~prediction strategy uses a
rolling average to generate the forecast: the prediction $p_i$ is
simply the average of [[all]] [[no window?  if so, then it's not a
    ``rolling'' average]] observed points up to that point: $$p_i =
\sum_{j=1}^{i-1}\frac{X_{j,obs}}{i-1}$$

While both of these methods are simplistic, they are not without
merit.  For a time series near the high end of the complexity
spectrum---i.e., one that possesses very little predictive
structure---these two methods can actually be the best choice.  In
forecasting currency exchange rates, for instance, sophisticated
econometric techniques and prediction models fail to consistently
outperform the random-walk method~\cite{rwMeese,rwCCE}.  These signals
are volatile, noisy, and possess very little predictive structure, but
they do not---on the average---vary a great deal, so guessing the last
known value is not a bad strategy.  If a signal has a unimodal
distribution with low variance, the \naive ~prediction strategy will
perform quite well---even if the signal is highly complex---simply
because the mean is a good approximation of the future behavior.
Also, a temporal average effects a low-pass filtering operation.  In a
signal with very little predictive structure, then, the \naive
~prediction strategy effectively mitigates the complexity.

Both of these methods have significant weaknesses, however, because
they do not model the temporal patterns of the time series or even the
distribution of its values.  Thus, they cannot track changes in that
structure, which causes them to fail in a number of important
situations.

Random-walk strategies are a particularly bad choice for time series
that change drastically at every time step.  In the worst case---a
large-amplitude square wave whose period is equivalent to twice the
sample time---a random-walk prediction would be exactly 180 degrees
out of phase with the true continuation.  The \naive ~method would be
better in this situation, since it would always predict the mean.  It
would, however, fail when a signal has a number of long-lived regimes
that have significantly different means.  In this situation, the
inertia of the \naive ~method's rolling mean is a liability and the
random-walk method's nimbleness is an advantage, allowing it to
react quickly to regime shifts.

Of course, a method that could capture and exploit periodic patterns
would be far more effective in these examples.  The LMA method
introduced in Section~\ref{sec:lma} is designed to do exactly that.
[[Couldn't one think about ARIMA as moving in that direction, too?]]
Note, though, that periodic patterns appear only in signals that are
at the low end of the complexity spectrum.  If a signal contains
little predictive structure, forecast strategies like LMA [[and
    ARIMA?]] have nothing to exploit and thus will be beaten out by
the two simple strategies described in this section.  This is explored
further in Sections~\ref{sec:accuracy} and~\ref{sec:results}

\subsection{Linear Forecasting: Seasonal Autoregressive-integrated-moving average(ARIMA){\color{blue} EDITABLE}}\label{sec:arima}
%\begin{enumerate}
%\item\cmark introduce method abstractly and practically, common method basically fitting a hyperplane to data removing seasonality and filtering for noise
%\item\cmark define rigorously, all the backshift operator garbage

@@

When there is evidence a generating process exhibits predictive structure an obvious strategy would be to fit a hyperplane to the dataset and then use this hyperplane to forecast new data points. This is a very old technique stemming back to Yule in 1927\cite{weigend93} with his invention of autoregressive forecasting, which forecasts the next time step through a weighted average of the past $p$ time-steps, i.e., $$p_i = \phi(B)X_{i,obs}.$$ Here $\phi(z)$ is a polynomial of degree $p$ and $B$ is the backshift operator, i.e., $$B^mX_i  = X_{i-m,obs}.$$
To improve this method further--especially in the presence of noisy data---one could allow for some wiggle room by adding a moving average of white noise to the model: $\theta(B)\epsilon_i$, where $\theta(z)$ is a polynomial of degree $q$ and $\{\epsilon_i\}\sim WN(0,\sigma^2)$. Finally if there is evidence of non-stationarity an initial differencing (integrated) is applied as an attempt to remove that non-stationarity: $(1-B^d)X_i$. Combining these three give us a nonseasonal  ARIMA$(p,d,q)$ model:  $$\phi(B)(1-B)^dX_i = \theta(B)\epsilon_i.$$ Where $p$, $d$ and $q$ correspond to the orders of the autoregressive, integrated and moving average models respectively. If periodic structure is present a seasonal ARIMA$(p,d,q)(P,D,Q)_m$ is often used and is defined by:$$\Phi(B^m)\phi(B)(1-B^m)^D(1-B)^dX_i = \Theta(B^m)\theta(B)\epsilon_i.$$ Here $\Phi(z)$ and $\Theta(z)$ are polynomials of degree P and Q respectively, and $m$ is the seasonal frequency. Forecasting is then done in two steps, first the integrated step is performed to remove non-stationarity. $$\hat{X_i} = (1-B^m)^D(1-B)^d X_{i,obs}.$$ Then forecasts are made by evaluating $$\Phi(B^m)\phi(B)\hat{X_i} = \Theta(B^m)\theta(B)\epsilon_i.$$

There is a vast amount of theory and literature on constructing, fitting and forecasting with this class of models. However, this is not the focus of this paper and  we refer the reader to \cite{davislinearts} for an in-depth exploration. In this paper, we intend to explore how predictive accuracy of several different forecasting schema correlates with signal complexity. As such, we use ARIMA$(p,d,q)(P,D,Q)_m$ (or simply ARIMA) models as an out of the box black box linear forecasting scheme. To do this we use an automated fitting and forecasting techniques described in \cite{autoARIMA}. This technique automatically selects model and differencing orders: $p, d, q, P, D, Q$ and $m$ as well as fitting the corresponding polynomials to the time series. Hyndman and Khandakar use unit root tests to select differencing orders, $d$ and $D$. In particular, they use a combination of  
KPSS unit-root tests \cite{KPSSunit} and their own extension of Canova-Hansen test \cite{Canova1995}. To do model order selection they use the Akaike information criterion (AIC) \cite{akaike1974} conditioned on the maximized likelihood of the  model fitted to the differenced data, $(1-B^m)^D(1-B)^d X_{i,obs}$ \cite{autoARIMA}. For more detail on this fitting process see \cite{autoARIMA}.

ARIMA forecasting is a time tested modelling process which uses a global linear fit and adjustments for seasonality, non-stationarity and noise. If a generating process fits into this category then this model is a great short term predictor. If information is instead being transmitted in a nonlinear way this global linear fit falls apart. One other weakness of this method is that if long prediction horizons are needed the forecast is guaranteed to converge to the mean after some number of predictions (dependent on model order). For this reason, we limit all methods to doing multiple 1-step predictions as to not handicap this method.

\subsection{Nonlinear Forecasting: Lorenz Method of Analogues (LMA){\color{blue} EDITABLE}}\label{sec:lma}

>>>>

In this section, we describe the four different forecasting schema
used in this study, as well as the error metric used to evaluate their
predictive accuracy.  These methods include:
\begin{itemize}
\item The \emph{random-walk} method, which uses the previous value in
  the observed signal as the forecast,

@@

\item The \emph{LMA} method, which uses a near-neighbor forecast
  strategy on a dynamical reconstruction of the signal.
\end{itemize}
ARIMA is based on standard autoregressive techniques.  LMA is designed
to capture and exploit the deterministic structure of a signal from a
nonlinear dynamical system.  The first two methods, somewhat
surprisingly, often outperform these more-sophisticated prediction
strategies in the case of highly complex signals, as discussed below.

@@

A random-walk predictor simply uses the last observed measurement as
the forecast: that is, the value $px_i$ at time $i$ is predicted to
be $$px_i = x_{i-1}$$ The \naive ~prediction strategy uses an average
to generate the forecast: the prediction $px_i$ is simply the running
average of observervations up to that point: $$px_i =
\sum_{j=1}^{i-1}\frac{x_j}{i-1}$$ While both of these methods are
simplistic, they are not without merit.  For a time series near the
high end of the complexity spectrum---i.e., one that possesses very
little predictive structure---these two methods can actually be the
best choice.  In forecasting currency exchange rates, for instance,
sophisticated econometric techniques and prediction models fail to
consistently outperform the random-walk method~\cite{rwMeese,rwCCE}.
These signals are volatile, noisy, and possess very little predictive
structure, but they do not---on the average---vary a great deal, so
the random-walk method's strategy of simply guessing the last known
value is not a bad choice.  If a signal has a unimodal distribution
with low variance, the \naive ~prediction strategy will perform quite
well---even if the signal is highly complex---simply because the mean
is a good approximation of the future behavior.  Also, a temporal
average effects a low-pass filtering operation.  In a signal with very
little predictive structure, then, the \naive ~prediction strategy
effectively mitigates the complexity.

Both of these methods have significant weaknesses, however, because
they do not model the temporal patterns of the time series or even the
distribution of its values.  Thus, they cannot track changes in that
structure, which causes them to fail in a number of important
situations.  Random-walk strategies are a particularly bad choice for
time series that change drastically at every time step.  In the worst
case---a large-amplitude square wave whose period is equivalent to
twice the sample time---a random-walk prediction would be exactly 180
degrees out of phase with the true continuation.  The \naive ~method
would be a better choice in this situation, since it would always
predict the mean.  It would, however, fail when a signal has a number
of long-lived regimes that have significantly different means.  In
this situation, the inertia of the \naive ~method's accumulating mean
is a liability and the agility of the random-walk method is an
advantage, since it can respond quickly to regime shifts.

Of course, methods that could capture and exploit the geometry of the
data, or its temporal patterns, would be far more effective in the
situations described in the previous paragraph.  The ARIMA and LMA
methods introduced in Sections~\ref{sec:lma} and~\ref{sec:arima} are
designed to do exactly that.  Note, though, that periodic patterns
appear only in signals that are at the low end of the complexity
spectrum.  If a signal contains little predictive structure, forecast
strategies like LMA and ARIMA have [[nothing to exploit and thus will
    be beaten out by]] [[needs wordsmithing from Liz]] the two simple
strategies described in this section.  This is explored further in
Sections~\ref{sec:accuracy} and~\ref{sec:results}.

\subsection{A Regression-Based Prediction Strategy}
\label{sec:arima}
%\begin{enumerate}
%\item\cmark introduce method abstractly and practically, common method basically fitting a hyperplane to data removing seasonality and filtering for noise
%\item\cmark define rigorously, all the backshift operator garbage

@@

Perhaps the simplest way to capture and exploit the structure of data
is to fit a hyperplane to the dataset and then use that it to forecast
new data points.  The roots of this date back to Yule's 1927 invention
of the autoregressive schema~\cite{weigend93}, which forecasts the
next time step through a weighted average of past observations: $$px_i
= \phi(B^p)x_{i}$$ where $\phi(\cdot)$ is a polynomial of degree $p$
that is fit to the past $p+1$ [[???]] values of $x_i$ via the
backshift operator $B$ \footnote{I changed the superscript to $k$ to
  make this definition general---not connected to any of the terms
  that have specific meanings later in this section.}:
$$B^k x_i = x_{i-k}$$
%
with $k=1...p$ [[presumably?]].  To account for noise in the data, one
can add a so-called ``moving average'' term to the model:
$\theta(B^q)\epsilon_i$, where $\theta(\cdot)$ is a polynomial of
degree $q$ and $\epsilon$ is white noise with the same mean and
variance as the data.
% if you leave this in, you'll need to explain a lot about the data
% and the calculation
% $\{\epsilon_i\}\sim WN(0,\sigma^2)$.  
To remove nonstationarities in the data, one can detrend it using a
differencing operation: $(1-B^d)x_i$.

A strategy that incorporates all three of these features is called a
\emph{nonseasonal ARIMA} model of order $(p,d,q)$:
 $$\phi(B^p)(1-B^d) x_i = \theta(B^q)\epsilon_i$$
%
{\color{red} You had superscripts on some of the $B$s and not on
  others.  I added them but I may have gotten them wrong.  I also
  changed some $(1-B)^d$s to $(1-B^d)$s to make things consistent with
  the first use of that construction.  Please make sure things are
  consistent, here and throughout this section.}  Here, $p$, $d$ and
$q$ correspond to the orders of the autoregressive, detrending, and
moving average terms, respectively.

If periodic structure is present in the data, a \emph{seasonal ARIMA}
model of order $(p,d,q)(P,D,Q)$ can be a good choice:
%
$$\Phi(B^P)\phi(B^p)(1-B^m)^D(1-B^d) x_i =
\Theta(B^Q)\theta(B^q)\epsilon_i$$
%
Here $\Phi(\cdot)$ and $\Theta(\cdot)$ are polynomials of degree $P$
and $Q$, respectively [[which accomplish what purpose?  why?  how?
    where did $D$ come from and what does it do?]].  $m$ is the
seasonal frequency [[shouldn't this be ``period''?]].  [[Explain why
    one should choose $P=Q=m$, since that's what the equation below
    implicitly does.]]  To use a seasonal ARIMA model, one first uses
the detrending term to remove nonstationarity: $$\hat{x_i} =
(1-B^m)^D(1-B^d) x_{i}$$ and then creates the forecast $px_i$ by
evaluating $$\Phi(B^m)\phi(B^p)\hat{X_i} =
\Theta(B^m)\theta(B^q)\epsilon_i$$

There is a vast amount of theory and literature regarding the
construction and use of this class of models; we refer the reader to
\cite{davislinearts} for an in-depth exploration.  For the purposes of
this paper, where the goal is to explore the relationship between
predictability and complexity across a broad array of forecast
strategies, seasonal ARIMA models are a good exemplar of the class of
linear forecasting strategies.  In the sections that follow, we use
the automated fitting and forecasting techniques described
in~\cite{autoARIMA} to select model and differencing orders and fit
the corresponding polynomials to the time series.  This procedure uses
a combination of KPSS unit-root tests \cite{KPSSunit} and a
customization of the Canova-Hansen test \cite{Canova1995} to select
the differencing orders $d$ and $D$ and chooses the model order using
the Akaike information criterion (AIC) \cite{akaike1974}, conditioned
on the maximized likelihood of the model fitted to the detrended data,
$(1-B^m)^D(1-B^d) x_{i,obs}$.

ARIMA forecasting is a common and time-tested procedure.  Its
adjustments for seasonality, nonstationarity, and noise make it an
appropriate choice for short-term predictions of time-series data
generated by a wide range of processes.  If information is being
generated and/or transmitted in a nonlinear way, however, a global
linear fit is inappropriate and ARIMA forecasts will be inaccurate.
Another weakness of this method is prediction horizon, since an ARIMA
forecast is guaranteed to converge to the mean after some number of
predictions, depending on model order.  To avoid this issue, we employ
it in a stepwise fashion: we build the model, perform a one-step
prediction, rebuild using the latest observations, and iterate until
the desired prediction horizon is reached.  (For consistency, we take
the same approach with the other three models in this study as well.)

\subsection{Nonlinear Forecasting: The Lorenz Method of Analogues}
\label{sec:lma}
